---
title: "Neurite and Glial Outgrowth Quantification"
author: "Mohapatra, Zhao, and Smirnova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  repo_root: "."
  outgrowth_root: "data/outgrowth"
  acute_metals_dir: "data/outgrowth/acute_metals"
  ml_validation_btubulin_dir: "data/outgrowth/ml_validation_btubulin"
  ml_validation_gfap_dir: "data/outgrowth/ml_validation_gfap"
  ml_vs_if_dir: "data/outgrowth/ml_vs_if"
  data_processed_dir: "data_processed"
  results_dir: "results"
  step_px: 10
  run_combine_sholl_profiles: false
  sholl_profile_parent_dir: "data/outgrowth/raw_sholl_profiles"
  save_png: true
  save_pdf: true
---

# Overview

This R Markdown workflow consolidates the neurite and glial outgrowth analysis used for the MortaLIVE brain organoid manuscript. The workflow starts from Fiji/Sholl-derived profiles or processed Sholl summary Excel files, converts the data to tidy long format, calculates area under the curve (AUC) for each sample, performs non-parametric statistical comparisons, and exports processed data tables and figures.

The workflow includes four modules:

1. Optional combination of Fiji-generated Sholl profile CSV files.
2. Acute metal exposure outgrowth analysis.
3. MortaLIVE validation analysis for beta-tubulin III and GFAP outgrowth.
4. MortaLIVE Live-channel versus immunofluorescence outgrowth comparisons.

The script assumes that image segmentation, thresholding, and Sholl profile generation have already been performed upstream in Fiji/ImageJ, Yokogawa CellPathfinder, or equivalent image-analysis software.

# Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "readxl",
  "readr",
  "dplyr",
  "tidyr",
  "tibble",
  "stringr",
  "purrr",
  "MESS",
  "ggplot2",
  "ggpubr",
  "rstatix",
  "writexl"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "The following required packages are not installed: ",
    paste(missing_packages, collapse = ", "),
    "\nInstall them before rendering this document."
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

repo_root <- normalizePath(params$repo_root, winslash = "/", mustWork = FALSE)

resolve_path <- function(path) {
  if (is.na(path) || !nzchar(path)) {
    return(NA_character_)
  }
  if (grepl("^([A-Za-z]:)?[/\\\\]", path)) {
    normalizePath(path, winslash = "/", mustWork = FALSE)
  } else {
    normalizePath(file.path(repo_root, path), winslash = "/", mustWork = FALSE)
  }
}

outgrowth_root <- resolve_path(params$outgrowth_root)
acute_metals_dir <- resolve_path(params$acute_metals_dir)
ml_validation_btubulin_dir <- resolve_path(params$ml_validation_btubulin_dir)
ml_validation_gfap_dir <- resolve_path(params$ml_validation_gfap_dir)
ml_vs_if_dir <- resolve_path(params$ml_vs_if_dir)
data_processed_dir <- resolve_path(params$data_processed_dir)
results_dir <- resolve_path(params$results_dir)
sholl_profile_parent_dir <- resolve_path(params$sholl_profile_parent_dir)

dir.create(data_processed_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

step_px <- params$step_px

message("Repository root: ", repo_root)
message("Processed data directory: ", data_processed_dir)
message("Results directory: ", results_dir)
```

# Helper functions

```{r helper-functions}
dir_has_files <- function(dir, pattern = "\\.xlsx$") {
  dir.exists(dir) && length(list.files(dir, pattern = pattern, full.names = TRUE, recursive = TRUE)) > 0
}

safe_file_stem <- function(x) {
  x |>
    stringr::str_replace_all("[^A-Za-z0-9_\\-]+", "_") |>
    stringr::str_replace_all("_+", "_") |>
    stringr::str_replace_all("^_|_$", "")
}

save_plot_outputs <- function(plot, filename_stem, width = 8, height = 5, dpi = 300) {
  if (isTRUE(params$save_pdf)) {
    ggplot2::ggsave(
      filename = file.path(results_dir, paste0(filename_stem, ".pdf")),
      plot = plot,
      width = width,
      height = height
    )
  }

  if (isTRUE(params$save_png)) {
    ggplot2::ggsave(
      filename = file.path(results_dir, paste0(filename_stem, ".png")),
      plot = plot,
      width = width,
      height = height,
      dpi = dpi
    )
  }
}

sem <- function(x) {
  stats::sd(x, na.rm = TRUE) / sqrt(sum(is.finite(x)))
}

p_to_stars <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ "ns"
  )
}

read_sholl_excel_long <- function(file, step_px = 10) {
  dat <- readxl::read_excel(file)

  if (!"Radius" %in% names(dat)) {
    stop("Input file does not contain a 'Radius' column: ", file)
  }

  dat |>
    dplyr::rename(edge_distance_px = Radius) |>
    dplyr::mutate(edge_distance_px = as.numeric(edge_distance_px)) |>
    dplyr::filter(is.finite(edge_distance_px), edge_distance_px %% step_px == 0) |>
    tidyr::pivot_longer(
      cols = -edge_distance_px,
      names_to = "Sample",
      values_to = "Inters"
    ) |>
    dplyr::mutate(
      SourceFile = basename(file),
      Sample = stringr::str_squish(as.character(Sample)),
      Inters = as.numeric(Inters)
    )
}

calculate_auc <- function(data, group_cols, x_col = "edge_distance_px", y_col = "Inters") {
  data |>
    dplyr::filter(is.finite(.data[[x_col]]), is.finite(.data[[y_col]])) |>
    dplyr::arrange(dplyr::across(dplyr::all_of(c(group_cols, x_col)))) |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_cols))) |>
    dplyr::summarise(
      n_points = dplyr::n(),
      AUC = dplyr::if_else(
        n_points >= 2,
        MESS::auc(.data[[x_col]], .data[[y_col]], type = "linear"),
        NA_real_
      ),
      x_min = min(.data[[x_col]], na.rm = TRUE),
      x_max = max(.data[[x_col]], na.rm = TRUE),
      .groups = "drop"
    )
}

pairwise_wilcox <- function(data, group_col, value_col = "AUC", reference_group = NULL, adjust_method = "BH") {
  dat <- data |>
    dplyr::filter(is.finite(.data[[value_col]]), !is.na(.data[[group_col]])) |>
    dplyr::mutate(.group = as.character(.data[[group_col]]))

  groups <- sort(unique(dat$.group))

  if (length(groups) < 2) {
    return(tibble::tibble())
  }

  if (!is.null(reference_group)) {
    pairs <- tibble::tibble(
      group1 = reference_group,
      group2 = setdiff(groups, reference_group)
    ) |>
      dplyr::filter(group2 != group1)
  } else {
    cmb <- t(utils::combn(groups, 2))
    pairs <- tibble::tibble(group1 = cmb[, 1], group2 = cmb[, 2])
  }

  if (nrow(pairs) == 0) {
    return(tibble::tibble())
  }

  res <- purrr::pmap_dfr(pairs, function(group1, group2) {
    x <- dat |> dplyr::filter(.group == group1) |> dplyr::pull(.data[[value_col]])
    y <- dat |> dplyr::filter(.group == group2) |> dplyr::pull(.data[[value_col]])

    if (sum(is.finite(x)) < 1 || sum(is.finite(y)) < 1) {
      return(tibble::tibble())
    }

    wt <- suppressWarnings(stats::wilcox.test(x, y, alternative = "two.sided", exact = FALSE))

    tibble::tibble(
      group1 = group1,
      group2 = group2,
      n1 = sum(is.finite(x)),
      n2 = sum(is.finite(y)),
      median1 = median(x, na.rm = TRUE),
      median2 = median(y, na.rm = TRUE),
      W = unname(wt$statistic),
      p = wt$p.value
    )
  })

  res |>
    dplyr::mutate(
      p_adj = stats::p.adjust(p, method = adjust_method),
      signif = p_to_stars(p_adj)
    )
}
```

# Optional preprocessing: combine Fiji Sholl CSV outputs

This optional section combines Fiji-generated `*_Sholl-Profiles.csv` files within each top-level condition folder into one Excel workbook per folder. Use this only when starting from raw Fiji/Sholl CSV outputs. If you already have processed Sholl summary Excel files, leave `run_combine_sholl_profiles` set to `false`.

```{r combine-sholl-profiles}
extract_sample_name <- function(file_path, folder_name) {
  file_name <- basename(file_path) |>
    stringr::str_remove("\\.tif_Sholl-Profiles\\.csv$") |>
    stringr::str_remove("\\.csv$")

  file_name <- stringr::str_remove(file_name, paste0("^", stringr::fixed(folder_name)))

  sample_name <- stringr::str_extract(file_name, "[A-Z]-\\d+_F\\d+")

  if (is.na(sample_name)) {
    sample_name <- file_name
  }

  sample_name
}

process_single_sholl_csv <- function(file_path, folder_name) {
  sample_name <- extract_sample_name(file_path, folder_name)

  read.csv(file_path, check.names = FALSE) |>
    dplyr::select(Radius, `Inters.`) |>
    dplyr::rename(!!sample_name := `Inters.`)
}

combine_one_sholl_folder <- function(folder_path) {
  folder_name <- basename(folder_path)

  file_list <- list.files(
    path = folder_path,
    pattern = "_Sholl-Profiles\\.csv$",
    full.names = TRUE,
    recursive = TRUE
  )

  if (length(file_list) == 0) {
    message("Skipped ", folder_name, ": no Sholl CSV files found.")
    return(NULL)
  }

  combined_table <- file_list |>
    purrr::map(~ process_single_sholl_csv(.x, folder_name)) |>
    purrr::reduce(dplyr::full_join, by = "Radius") |>
    dplyr::arrange(Radius)

  output_file <- file.path(folder_path, paste0(folder_name, "_Sholl_Combined.xlsx"))

  writexl::write_xlsx(
    x = list(Sholl = combined_table),
    path = output_file
  )

  message("Saved: ", output_file)
  combined_table
}

if (isTRUE(params$run_combine_sholl_profiles)) {
  if (!dir.exists(sholl_profile_parent_dir)) {
    stop("Sholl profile parent directory does not exist: ", sholl_profile_parent_dir)
  }

  top_level_folders <- list.dirs(sholl_profile_parent_dir, recursive = FALSE, full.names = TRUE)

  combined_sholl_results <- top_level_folders |>
    stats::setNames(basename(top_level_folders)) |>
    purrr::map(combine_one_sholl_folder)
} else {
  message("Skipping raw Sholl CSV combination.")
}
```

# Module 1: Acute metal exposure outgrowth analysis

```{r acute-metals-functions}
parse_acute_metals <- function(data) {
  parsed <- data |>
    dplyr::mutate(
      is_ctrl = stringr::str_detect(Sample, stringr::regex("^(ctrl|control)", ignore_case = TRUE)),
      Dose_uM = dplyr::if_else(
        is_ctrl,
        0,
        as.numeric(stringr::str_match(Sample, "^([0-9.]+)([A-Za-z]+)_?([0-9]+)$")[, 2])
      ),
      Metal = dplyr::if_else(
        is_ctrl,
        "Ctrl",
        stringr::str_match(Sample, "^([0-9.]+)([A-Za-z]+)_?([0-9]+)$")[, 3]
      ),
      Rep = dplyr::if_else(
        is_ctrl,
        stringr::str_extract(Sample, "[0-9]+$"),
        stringr::str_match(Sample, "^([0-9.]+)([A-Za-z]+)_?([0-9]+)$")[, 4]
      ),
      Metal = dplyr::case_when(
        stringr::str_to_lower(Metal) %in% c("crvi", "cr") ~ "Cr",
        stringr::str_to_lower(Metal) %in% c("cd") ~ "Cd",
        stringr::str_to_lower(Metal) %in% c("pb") ~ "Pb",
        stringr::str_to_lower(Metal) %in% c("mix", "cdcr", "cdcrvi", "cdcrmix") ~ "Mix",
        TRUE ~ Metal
      ),
      Condition = dplyr::if_else(Metal == "Ctrl", "Ctrl", paste0(Metal, "_", Dose_uM))
    ) |>
    dplyr::select(SourceFile, Sample, is_ctrl, Metal, Dose_uM, Rep, Condition, edge_distance_px, Inters)

  unparsable <- parsed |>
    dplyr::filter(!is_ctrl & (is.na(Dose_uM) | is.na(Metal) | is.na(Rep))) |>
    dplyr::distinct(SourceFile, Sample)

  if (nrow(unparsable) > 0) {
    warning(
      "Some acute metal sample names could not be parsed. Review these names:\n",
      paste(utils::capture.output(print(unparsable, n = Inf)), collapse = "\n")
    )
  }

  parsed |>
    dplyr::filter(!is.na(Dose_uM), !is.na(Metal), !is.na(Rep)) |>
    dplyr::select(-is_ctrl)
}

run_pairwise_wilcox_by_metal <- function(df_auc_curve,
                                         metal_col = "Metal",
                                         dose_col = "Dose_uM",
                                         value_col = "AUC",
                                         adjust_method = "BH",
                                         low_n_cutoff = 3) {
  dat <- df_auc_curve |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::mutate(!!dose_col := as.numeric(.data[[dose_col]]))

  metals <- sort(unique(dat[[metal_col]]))

  purrr::map_dfr(metals, function(m) {
    df_m <- dat |> dplyr::filter(.data[[metal_col]] == m)
    doses <- sort(unique(df_m[[dose_col]]))

    if (length(doses) < 2) {
      return(tibble::tibble())
    }

    pairs <- t(utils::combn(doses, 2))

    res_m <- purrr::map_dfr(seq_len(nrow(pairs)), function(i) {
      d1 <- pairs[i, 1]
      d2 <- pairs[i, 2]

      x <- df_m |> dplyr::filter(.data[[dose_col]] == d1) |> dplyr::pull(.data[[value_col]])
      y <- df_m |> dplyr::filter(.data[[dose_col]] == d2) |> dplyr::pull(.data[[value_col]])

      n1 <- sum(is.finite(x))
      n2 <- sum(is.finite(y))

      if (n1 < 1 || n2 < 1) {
        return(tibble::tibble())
      }

      wt <- suppressWarnings(stats::wilcox.test(x, y, alternative = "two.sided", exact = FALSE))

      tibble::tibble(
        Metal = m,
        dose1 = d1,
        dose2 = d2,
        n1 = n1,
        n2 = n2,
        median1 = median(x, na.rm = TRUE),
        median2 = median(y, na.rm = TRUE),
        W = unname(wt$statistic),
        p = wt$p.value,
        low_n_flag = (n1 < low_n_cutoff || n2 < low_n_cutoff)
      )
    })

    res_m |> dplyr::mutate(p_adj = stats::p.adjust(p, method = adjust_method))
  }) |>
    dplyr::mutate(signif = p_to_stars(p_adj))
}
```

```{r acute-metals-analysis}
if (dir_has_files(acute_metals_dir, "\\.xlsx$")) {
  acute_files <- list.files(acute_metals_dir, pattern = "\\.xlsx$", full.names = TRUE, recursive = TRUE)

  acute_long <- acute_files |>
    purrr::map_dfr(~ read_sholl_excel_long(.x, step_px = step_px)) |>
    parse_acute_metals()

  readr::write_csv(acute_long, file.path(data_processed_dir, "acute_metals_long_table.csv"))

  acute_nonctrl_curve_summary <- acute_long |>
    dplyr::filter(Metal != "Ctrl") |>
    dplyr::group_by(Metal, Dose_uM, edge_distance_px) |>
    dplyr::summarise(
      mean_inters = mean(Inters, na.rm = TRUE),
      sem_inters = sem(Inters),
      n = sum(is.finite(Inters)),
      .groups = "drop"
    )

  acute_ctrl_curve_summary <- acute_long |>
    dplyr::filter(Metal == "Ctrl", Dose_uM == 0) |>
    dplyr::group_by(edge_distance_px) |>
    dplyr::summarise(
      mean_inters = mean(Inters, na.rm = TRUE),
      sem_inters = sem(Inters),
      n = sum(is.finite(Inters)),
      Dose_uM = 0,
      .groups = "drop"
    )

  acute_metals <- acute_nonctrl_curve_summary |>
    dplyr::distinct(Metal) |>
    dplyr::pull(Metal) |>
    sort()

  acute_curve_summary <- dplyr::bind_rows(
    acute_nonctrl_curve_summary,
    tidyr::crossing(acute_ctrl_curve_summary, Metal = acute_metals)
  ) |>
    dplyr::mutate(
      Dose_f = factor(
        as.numeric(Dose_uM),
        levels = c(0, 0.32, 3.2, 10, 50),
        labels = c("0", "0.32", "3.2", "10", "50")
      ),
      Metal = factor(Metal, levels = intersect(c("Cd", "Cr", "Mix", "Pb"), unique(Metal)))
    )

  readr::write_csv(acute_curve_summary, file.path(data_processed_dir, "acute_metals_curve_summary.csv"))

  p_acute_curve <- ggplot2::ggplot(
    acute_curve_summary,
    ggplot2::aes(x = edge_distance_px, y = mean_inters, color = Dose_f, fill = Dose_f)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_ribbon(
      ggplot2::aes(ymin = mean_inters - sem_inters, ymax = mean_inters + sem_inters),
      alpha = 0.18,
      color = NA
    ) +
    ggplot2::facet_wrap(~Metal, nrow = 1, scales = "free_y") +
    ggplot2::labs(
      title = "Acute metal exposure outgrowth curves",
      x = "Distance from organoid edge (pixel)",
      y = "Intersections",
      color = "Exposure (uM)",
      fill = "Exposure (uM)"
    ) +
    ggplot2::theme_classic()

  print(p_acute_curve)
  save_plot_outputs(p_acute_curve, "acute_metals_curve", width = 12, height = 4)

  acute_auc_curve <- calculate_auc(
    acute_long,
    group_cols = c("Metal", "Dose_uM", "Rep", "Condition")
  )

  acute_auc_summary <- acute_auc_curve |>
    dplyr::group_by(Metal, Dose_uM) |>
    dplyr::summarise(
      mean_AUC = mean(AUC, na.rm = TRUE),
      sem_AUC = sem(AUC),
      n_rep = sum(is.finite(AUC)),
      .groups = "drop"
    )

  readr::write_csv(acute_auc_curve, file.path(data_processed_dir, "acute_metals_auc_per_curve.csv"))
  readr::write_csv(acute_auc_summary, file.path(data_processed_dir, "acute_metals_auc_summary.csv"))

  acute_ctrl_auc <- acute_auc_curve |>
    dplyr::filter(Metal == "Ctrl", Dose_uM == 0) |>
    dplyr::select(-Metal)

  acute_auc_for_test <- dplyr::bind_rows(
    acute_auc_curve |> dplyr::filter(Metal != "Ctrl"),
    tidyr::crossing(acute_ctrl_auc, Metal = acute_metals)
  )

  acute_pairwise_table <- run_pairwise_wilcox_by_metal(acute_auc_for_test)
  readr::write_csv(acute_pairwise_table, file.path(data_processed_dir, "acute_metals_auc_pairwise_wilcox.csv"))

  acute_normality <- acute_auc_for_test |>
    dplyr::group_by(Metal, Dose_uM) |>
    dplyr::summarise(
      n = sum(is.finite(AUC)),
      shapiro_p = {
        vals <- AUC[is.finite(AUC)]
        if (length(vals) >= 3) stats::shapiro.test(vals)$p.value else NA_real_
      },
      .groups = "drop"
    )

  readr::write_csv(acute_normality, file.path(data_processed_dir, "acute_metals_auc_normality.csv"))

  acute_auc_plot_data <- acute_auc_for_test |>
    dplyr::mutate(
      Dose_uM = as.numeric(Dose_uM),
      Dose_f = factor(
        Dose_uM,
        levels = c(0, 0.32, 3.2, 10, 50),
        labels = c("0", "0.32", "3.2", "10", "50")
      ),
      Metal = factor(Metal, levels = intersect(c("Cd", "Cr", "Mix", "Pb"), unique(Metal)))
    )

  acute_anno <- acute_pairwise_table |>
    dplyr::filter((dose1 == 0 | dose2 == 0), p_adj < 0.05) |>
    dplyr::mutate(
      group1 = factor(dose1, levels = c(0, 0.32, 3.2, 10, 50), labels = c("0", "0.32", "3.2", "10", "50")),
      group2 = factor(dose2, levels = c(0, 0.32, 3.2, 10, 50), labels = c("0", "0.32", "3.2", "10", "50")),
      label = signif
    )

  if (nrow(acute_anno) > 0) {
    y_base <- acute_auc_plot_data |>
      dplyr::group_by(Metal) |>
      dplyr::summarise(y0 = max(AUC, na.rm = TRUE), .groups = "drop")

    acute_anno <- acute_anno |>
      dplyr::left_join(y_base, by = "Metal") |>
      dplyr::group_by(Metal) |>
      dplyr::arrange(p_adj, .by_group = TRUE) |>
      dplyr::mutate(y.position = y0 * (1.08 + 0.08 * (dplyr::row_number() - 1))) |>
      dplyr::ungroup()
  }

  p_acute_auc <- ggplot2::ggplot(acute_auc_plot_data, ggplot2::aes(x = Dose_f, y = AUC, fill = Dose_f)) +
    ggplot2::stat_summary(fun = mean, geom = "col", width = 0.65, alpha = 0.9) +
    ggplot2::stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.16, linewidth = 0.45) +
    ggplot2::geom_jitter(width = 0.08, size = 2.6, shape = 21, color = "black", stroke = 0.3, alpha = 0.9) +
    ggplot2::facet_wrap(~Metal, nrow = 1, scales = "free_y") +
    ggplot2::labs(
      title = "Acute metal exposure outgrowth AUC",
      x = "Exposure (uM)",
      y = "AUC"
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "none")

  if (nrow(acute_anno) > 0) {
    p_acute_auc <- p_acute_auc +
      ggpubr::stat_pvalue_manual(
        data = acute_anno,
        label = "label",
        xmin = "group1",
        xmax = "group2",
        y.position = "y.position",
        tip.length = 0.01,
        size = 5,
        inherit.aes = FALSE
      )
  }

  print(p_acute_auc)
  save_plot_outputs(p_acute_auc, "acute_metals_auc_barplot", width = 12, height = 5)

} else {
  message("Skipping acute metal exposure analysis because no Excel files were found in: ", acute_metals_dir)
}
```

# Module 2: MortaLIVE validation analysis

This section analyzes the beta-tubulin III and GFAP validation datasets using the same workflow. Each dataset is converted to long format, summarized as an outgrowth curve, converted to replicate-level AUC values, normalized to the control mean, and compared against the control group using Wilcoxon rank-sum tests.

```{r ml-validation-functions}
parse_ml_validation <- function(data) {
  data |>
    dplyr::mutate(
      Group = dplyr::case_when(
        stringr::str_detect(Sample, stringr::regex("^(ctrl|control)", ignore_case = TRUE)) ~ "Ctrl",
        stringr::str_detect(Sample, "^0\\.0005ML_\\d+$") ~ "0.0005ML",
        stringr::str_detect(Sample, "^0\\.001ML_\\d+$") ~ "0.001ML",
        TRUE ~ NA_character_
      ),
      Rep = stringr::str_extract(Sample, "(?<=_)\\d+$"),
      Rep = stringr::str_pad(Rep, width = 2, side = "left", pad = "0"),
      Group = factor(Group, levels = c("Ctrl", "0.0005ML", "0.001ML"))
    ) |>
    dplyr::filter(!is.na(Group)) |>
    dplyr::select(SourceFile, Sample, Group, Rep, edge_distance_px, Inters)
}

run_ml_validation <- function(input_dir, assay_label, output_prefix) {
  if (!dir_has_files(input_dir, "\\.xlsx$")) {
    message("Skipping ", assay_label, " validation because no Excel files were found in: ", input_dir)
    return(invisible(NULL))
  }

  files <- list.files(input_dir, pattern = "\\.xlsx$", full.names = TRUE, recursive = TRUE)

  df_long <- files |>
    purrr::map_dfr(~ read_sholl_excel_long(.x, step_px = step_px)) |>
    parse_ml_validation()

  readr::write_csv(df_long, file.path(data_processed_dir, paste0(output_prefix, "_long_table.csv")))

  df_curve_summary <- df_long |>
    dplyr::group_by(Group, edge_distance_px) |>
    dplyr::summarise(
      mean_inters = mean(Inters, na.rm = TRUE),
      sem_inters = sem(Inters),
      n = sum(is.finite(Inters)),
      .groups = "drop"
    )

  readr::write_csv(df_curve_summary, file.path(data_processed_dir, paste0(output_prefix, "_curve_summary.csv")))

  p_curve <- ggplot2::ggplot(df_curve_summary, ggplot2::aes(edge_distance_px, mean_inters, color = Group, fill = Group)) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_ribbon(
      ggplot2::aes(ymin = mean_inters - sem_inters, ymax = mean_inters + sem_inters),
      alpha = 0.18,
      color = NA
    ) +
    ggplot2::labs(
      title = paste0("MortaLIVE validation: ", assay_label, " outgrowth curve"),
      x = "Distance from organoid edge (pixel)",
      y = "Intersections",
      color = "Group",
      fill = "Group"
    ) +
    ggplot2::theme_classic()

  print(p_curve)
  save_plot_outputs(p_curve, paste0(output_prefix, "_curve"), width = 7, height = 5)

  df_auc <- calculate_auc(df_long, group_cols = c("Group", "Rep", "Sample"))

  readr::write_csv(df_auc, file.path(data_processed_dir, paste0(output_prefix, "_auc_per_curve.csv")))

  df_auc_summary <- df_auc |>
    dplyr::group_by(Group) |>
    dplyr::summarise(
      mean_AUC = mean(AUC, na.rm = TRUE),
      sem_AUC = sem(AUC),
      n = sum(is.finite(AUC)),
      .groups = "drop"
    )

  readr::write_csv(df_auc_summary, file.path(data_processed_dir, paste0(output_prefix, "_auc_summary.csv")))

  df_pairwise <- pairwise_wilcox(
    data = df_auc,
    group_col = "Group",
    value_col = "AUC",
    reference_group = "Ctrl"
  )

  readr::write_csv(df_pairwise, file.path(data_processed_dir, paste0(output_prefix, "_auc_pairwise_wilcox.csv")))

  ctrl_mean_auc <- df_auc |>
    dplyr::filter(Group == "Ctrl", is.finite(AUC)) |>
    dplyr::summarise(m = mean(AUC, na.rm = TRUE)) |>
    dplyr::pull(m)

  df_auc_norm <- df_auc |>
    dplyr::filter(is.finite(AUC)) |>
    dplyr::mutate(
      AUC_norm = AUC / ctrl_mean_auc,
      log2FC = log2(AUC / ctrl_mean_auc)
    )

  readr::write_csv(df_auc_norm, file.path(data_processed_dir, paste0(output_prefix, "_auc_normalized.csv")))

  p_auc <- ggplot2::ggplot(df_auc, ggplot2::aes(x = Group, y = AUC, fill = Group)) +
    ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.65) +
    ggplot2::geom_jitter(width = 0.12, size = 2.3, alpha = 0.9, shape = 21, color = "black", stroke = 0.25) +
    ggplot2::labs(
      title = paste0("MortaLIVE validation: ", assay_label, " outgrowth AUC"),
      x = NULL,
      y = "AUC"
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "none")

  print(p_auc)
  save_plot_outputs(p_auc, paste0(output_prefix, "_auc"), width = 6, height = 5)

  p_norm <- ggplot2::ggplot(df_auc_norm, ggplot2::aes(x = Group, y = AUC_norm, fill = Group)) +
    ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.65) +
    ggplot2::geom_jitter(width = 0.12, size = 2.3, alpha = 0.9, shape = 21, color = "black", stroke = 0.25) +
    ggplot2::geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.4) +
    ggplot2::labs(
      title = paste0("MortaLIVE validation: ", assay_label, " AUC normalized to control mean"),
      x = NULL,
      y = "Control-normalized AUC"
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "none")

  print(p_norm)
  save_plot_outputs(p_norm, paste0(output_prefix, "_auc_normalized"), width = 6, height = 5)

  invisible(
    list(
      long = df_long,
      curve_summary = df_curve_summary,
      auc = df_auc,
      auc_normalized = df_auc_norm,
      pairwise = df_pairwise
    )
  )
}
```

```{r ml-validation-analysis}
ml_btubulin_results <- run_ml_validation(
  input_dir = ml_validation_btubulin_dir,
  assay_label = "beta-tubulin III",
  output_prefix = "ml_validation_btubulin"
)

ml_gfap_results <- run_ml_validation(
  input_dir = ml_validation_gfap_dir,
  assay_label = "GFAP",
  output_prefix = "ml_validation_gfap"
)
```

# Module 3: MortaLIVE Live-channel versus immunofluorescence comparison

## Two-group ML versus IF comparison

```{r ml-vs-if-two-group}
parse_ml_vs_if_two_group <- function(data) {
  data |>
    dplyr::mutate(
      Method = dplyr::case_when(
        stringr::str_detect(Sample, "_IF_") ~ "IF",
        stringr::str_detect(Sample, "_ML_") ~ "ML",
        TRUE ~ NA_character_
      ),
      Condition = stringr::str_extract(Sample, "^\\d+\\.\\d+ML"),
      PairID = stringr::str_extract(Sample, "(?<=_)\\d+$"),
      PairID = stringr::str_pad(PairID, width = 2, side = "left", pad = "0"),
      Method = factor(Method, levels = c("IF", "ML"))
    ) |>
    dplyr::filter(!is.na(Method), !is.na(PairID)) |>
    dplyr::select(SourceFile, Sample, Method, Condition, PairID, edge_distance_px, Inters)
}

if (dir_has_files(ml_vs_if_dir, "\\.xlsx$")) {
  ml_vs_if_files <- list.files(ml_vs_if_dir, pattern = "\\.xlsx$", full.names = TRUE, recursive = TRUE)

  ml_vs_if_long <- ml_vs_if_files |>
    purrr::map_dfr(~ read_sholl_excel_long(.x, step_px = step_px)) |>
    parse_ml_vs_if_two_group()

  if (nrow(ml_vs_if_long) > 0) {
    readr::write_csv(ml_vs_if_long, file.path(data_processed_dir, "ml_vs_if_two_group_long_table.csv"))

    ml_vs_if_curve_summary <- ml_vs_if_long |>
      dplyr::group_by(Method, edge_distance_px) |>
      dplyr::summarise(
        mean_inters = mean(Inters, na.rm = TRUE),
        sem_inters = sem(Inters),
        n = sum(is.finite(Inters)),
        .groups = "drop"
      )

    readr::write_csv(ml_vs_if_curve_summary, file.path(data_processed_dir, "ml_vs_if_two_group_curve_summary.csv"))

    p_ml_vs_if_curve <- ggplot2::ggplot(
      ml_vs_if_curve_summary,
      ggplot2::aes(edge_distance_px, mean_inters, color = Method, fill = Method)
    ) +
      ggplot2::geom_line(linewidth = 1) +
      ggplot2::geom_ribbon(
        ggplot2::aes(ymin = mean_inters - sem_inters, ymax = mean_inters + sem_inters),
        alpha = 0.18,
        color = NA
      ) +
      ggplot2::labs(
        title = "Outgrowth curve: IF versus MortaLIVE Live-channel",
        x = "Distance from organoid edge (pixel)",
        y = "Intersections",
        color = "Method",
        fill = "Method"
      ) +
      ggplot2::theme_classic()

    print(p_ml_vs_if_curve)
    save_plot_outputs(p_ml_vs_if_curve, "ml_vs_if_two_group_curve", width = 7, height = 5)

    ml_vs_if_auc <- calculate_auc(ml_vs_if_long, group_cols = c("Method", "PairID", "Sample"))

    readr::write_csv(ml_vs_if_auc, file.path(data_processed_dir, "ml_vs_if_two_group_auc_per_curve.csv"))

    ml_vs_if_stats <- pairwise_wilcox(
      data = ml_vs_if_auc,
      group_col = "Method",
      value_col = "AUC",
      reference_group = "IF"
    )

    readr::write_csv(ml_vs_if_stats, file.path(data_processed_dir, "ml_vs_if_two_group_wilcox.csv"))

    p_text <- ifelse(nrow(ml_vs_if_stats) > 0, signif(ml_vs_if_stats$p_adj[1], 3), NA)

    p_ml_vs_if_auc <- ggplot2::ggplot(ml_vs_if_auc, ggplot2::aes(x = Method, y = AUC, fill = Method)) +
      ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.65) +
      ggplot2::geom_jitter(width = 0.12, size = 2.3, alpha = 0.9, shape = 21, color = "black", stroke = 0.25) +
      ggplot2::labs(
        title = paste0("Outgrowth AUC: IF versus ML (Wilcoxon BH-adjusted p = ", p_text, ")"),
        x = NULL,
        y = "AUC"
      ) +
      ggplot2::theme_classic() +
      ggplot2::theme(legend.position = "none")

    print(p_ml_vs_if_auc)
    save_plot_outputs(p_ml_vs_if_auc, "ml_vs_if_two_group_auc", width = 5, height = 5)
  } else {
    message("Skipping two-group ML versus IF comparison because no samples matched the expected '_IF_' or '_ML_' naming pattern.")
  }
} else {
  message("Skipping ML versus IF comparison because no Excel files were found in: ", ml_vs_if_dir)
}
```

## Three-group ML, beta-tubulin III IF, and GFAP IF comparison

This section expects three files in `data/outgrowth/ml_vs_if/`:

```text
IF.xlsx
IF_GFAP.xlsx
ML.xlsx
```

```{r ml-vs-if-three-group}
parse_ml_vs_if_three_group_file <- function(file) {
  method_from_file <- dplyr::case_when(
    stringr::str_to_lower(basename(file)) == "if.xlsx" ~ "IF_bTubIII",
    stringr::str_to_lower(basename(file)) == "if_gfap.xlsx" ~ "IF_GFAP",
    stringr::str_to_lower(basename(file)) == "ml.xlsx" ~ "ML",
    TRUE ~ NA_character_
  )

  if (is.na(method_from_file)) {
    return(tibble::tibble())
  }

  read_sholl_excel_long(file, step_px = step_px) |>
    dplyr::mutate(
      Method = method_from_file,
      Rep = stringr::str_extract(Sample, "(?<=_)\\d+$"),
      Rep = stringr::str_pad(Rep, width = 2, side = "left", pad = "0"),
      Method = factor(Method, levels = c("IF_bTubIII", "IF_GFAP", "ML"))
    ) |>
    dplyr::filter(!is.na(Method)) |>
    dplyr::select(SourceFile, Sample, Method, Rep, edge_distance_px, Inters)
}

three_group_files <- file.path(ml_vs_if_dir, c("IF.xlsx", "IF_GFAP.xlsx", "ML.xlsx"))

if (all(file.exists(three_group_files))) {
  three_group_long <- three_group_files |>
    purrr::map_dfr(parse_ml_vs_if_three_group_file)

  readr::write_csv(three_group_long, file.path(data_processed_dir, "ml_vs_if_three_group_long_table.csv"))

  three_group_curve_summary <- three_group_long |>
    dplyr::group_by(Method, edge_distance_px) |>
    dplyr::summarise(
      mean_inters = mean(Inters, na.rm = TRUE),
      sem_inters = sem(Inters),
      n = sum(is.finite(Inters)),
      .groups = "drop"
    )

  readr::write_csv(three_group_curve_summary, file.path(data_processed_dir, "ml_vs_if_three_group_curve_summary.csv"))

  p_three_group_curve <- ggplot2::ggplot(
    three_group_curve_summary,
    ggplot2::aes(edge_distance_px, mean_inters, color = Method, fill = Method)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_ribbon(
      ggplot2::aes(ymin = mean_inters - sem_inters, ymax = mean_inters + sem_inters),
      alpha = 0.18,
      color = NA
    ) +
    ggplot2::labs(
      title = "Outgrowth curve: ML, beta-tubulin III IF, and GFAP IF",
      x = "Distance from organoid edge (pixel)",
      y = "Intersections",
      color = NULL,
      fill = NULL
    ) +
    ggplot2::theme_classic()

  print(p_three_group_curve)
  save_plot_outputs(p_three_group_curve, "ml_vs_if_three_group_curve", width = 7, height = 5)

  three_group_auc <- calculate_auc(three_group_long, group_cols = c("Method", "Sample"))

  three_group_auc <- three_group_auc |>
    dplyr::filter(is.finite(AUC)) |>
    dplyr::mutate(Method = factor(Method, levels = c("IF_bTubIII", "IF_GFAP", "ML")))

  readr::write_csv(three_group_auc, file.path(data_processed_dir, "ml_vs_if_three_group_auc_per_curve.csv"))

  three_group_stats <- dplyr::bind_rows(
    pairwise_wilcox(
      data = three_group_auc |> dplyr::filter(Method %in% c("IF_bTubIII", "ML")),
      group_col = "Method",
      value_col = "AUC",
      reference_group = "IF_bTubIII"
    ) |> dplyr::mutate(comparison = "IF_bTubIII_vs_ML"),
    pairwise_wilcox(
      data = three_group_auc |> dplyr::filter(Method %in% c("IF_GFAP", "ML")),
      group_col = "Method",
      value_col = "AUC",
      reference_group = "IF_GFAP"
    ) |> dplyr::mutate(comparison = "IF_GFAP_vs_ML")
  )

  readr::write_csv(three_group_stats, file.path(data_processed_dir, "ml_vs_if_three_group_wilcox.csv"))

  y_max <- max(three_group_auc$AUC, na.rm = TRUE)

  three_group_anno <- three_group_stats |>
    dplyr::transmute(
      group1 = group1,
      group2 = group2,
      label = signif,
      y.position = y_max * (1.08 + 0.08 * (dplyr::row_number() - 1))
    )

  p_three_group_auc <- ggplot2::ggplot(three_group_auc, ggplot2::aes(x = Method, y = AUC, fill = Method)) +
    ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.65) +
    ggplot2::geom_jitter(width = 0.12, size = 2.6, alpha = 0.9, shape = 21, color = "black", stroke = 0.25) +
    ggplot2::labs(
      title = "Outgrowth AUC: ML, beta-tubulin III IF, and GFAP IF",
      x = NULL,
      y = "AUC"
    ) +
    ggplot2::theme_classic(base_size = 16) +
    ggplot2::theme(legend.position = "none")

  if (nrow(three_group_anno) > 0) {
    p_three_group_auc <- p_three_group_auc +
      ggpubr::stat_pvalue_manual(
        data = three_group_anno,
        label = "label",
        xmin = "group1",
        xmax = "group2",
        y.position = "y.position",
        tip.length = 0.01,
        size = 5,
        inherit.aes = FALSE
      )
  }

  print(p_three_group_auc)
  save_plot_outputs(p_three_group_auc, "ml_vs_if_three_group_auc", width = 6, height = 5)
} else {
  missing_three_group_files <- three_group_files[!file.exists(three_group_files)]
  message(
    "Skipping three-group ML/IF/GFAP comparison. Missing expected file(s): ",
    paste(basename(missing_three_group_files), collapse = ", ")
  )
}
```

# Session information

```{r session-info}
sessionInfo()
```
