---
title: "Bliss Independence Additivity Analysis"
author: "Ronit Mohapatra, Hanyu Zhao, Lena Smirnova"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
params:
  input_path: "data/bliss/Quantification Table.xlsx"
  input_sheet: "RME quantification"
  output_dir: "data_processed"
  results_dir: "results"
  output_excel: "Bliss_Independence_Mixture_Effects.xlsx"
  n_boot: 5000
  seed: 831
  values_are_percent: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Purpose

This R Markdown workflow evaluates whether the observed Cd + Cr(VI) mixture effect is consistent with Bliss independence additivity. The expected Bliss effect is calculated from the single-agent Cd and Cr(VI) effects at each matched concentration:

\[
E_{Bliss} = E_{Cd} + E_{Cr} - (E_{Cd} \times E_{Cr})
\]

The Bliss excess is then calculated as:

\[
E_{excess} = E_{Mix,observed} - E_{Bliss}
\]

A non-parametric bootstrap is used to estimate a 95% confidence interval for the Bliss excess at each concentration. Mixture effects are classified as follows:

- **Greater than additive / synergistic**: 95% CI entirely above 0
- **Less than additive / antagonistic**: 95% CI entirely below 0
- **Additive**: 95% CI overlaps 0

# Required packages

```{r packages}
required_packages <- c(
  "readxl",
  "dplyr",
  "tidyr",
  "tibble",
  "stringr",
  "janitor",
  "readr",
  "writexl",
  "ggplot2"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    "The following required package(s) are not installed: ",
    paste(missing_packages, collapse = ", "),
    ". Install them before rendering this R Markdown file."
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

# User-defined paths and settings

This script is written to run from the repository root, following the structure described in the README.

```{r paths}
input_path <- params$input_path
input_sheet <- params$input_sheet
output_dir <- params$output_dir
results_dir <- params$results_dir
output_excel <- file.path(results_dir, params$output_excel)
n_boot <- params$n_boot
seed <- params$seed
values_are_percent <- params$values_are_percent

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(input_path)) {
  stop(
    "Input file not found: ", input_path,
    "\nUpdate params$input_path in the YAML header or render with a custom input_path parameter."
  )
}
```

# Read and validate input data

The expected input is a wide-format Excel sheet with one concentration column and replicate columns for Cd, Cr(VI), and the Cd + Cr(VI) mixture. After column-name cleaning, treatment replicate columns should follow this pattern:

```text
cd_1, cd_2, cr_1, cr_2, mix_1, mix_2, ...
```

The concentration column should be named `dose` before or after cleaning.

```{r read-data}
raw <- readxl::read_excel(
  path = input_path,
  sheet = input_sheet
) %>%
  janitor::clean_names()

if (!"dose" %in% names(raw)) {
  stop(
    "The input sheet must contain a concentration column named 'dose'. ",
    "Column names after cleaning are: ", paste(names(raw), collapse = ", ")
  )
}

if (ncol(raw) < 4) {
  stop("The input table must include 'dose' plus replicate columns for cd, cr, and mix.")
}

raw
```

# Convert to replicate-level long format

```{r tidy-data}
long <- raw %>%
  tidyr::pivot_longer(
    cols = -dose,
    names_to = c("treatment", "replicate"),
    names_pattern = "^([A-Za-z]+)_?(\\d+)$",
    values_to = "normalized_ld_input"
  ) %>%
  dplyr::mutate(
    treatment = stringr::str_to_lower(treatment),
    replicate = as.integer(replicate),
    normalized_ld_ratio = dplyr::if_else(
      values_are_percent,
      as.numeric(normalized_ld_input) / 100,
      as.numeric(normalized_ld_input)
    ),
    cytotoxicity_effect = 1 - normalized_ld_ratio
  ) %>%
  dplyr::select(
    dose,
    treatment,
    replicate,
    normalized_ld_input,
    normalized_ld_ratio,
    cytotoxicity_effect
  ) %>%
  dplyr::filter(!is.na(treatment), !is.na(replicate))

required_treatments <- c("cd", "cr", "mix")
observed_treatments <- sort(unique(long$treatment))
missing_treatments <- setdiff(required_treatments, observed_treatments)

if (length(missing_treatments) > 0) {
  stop(
    "Missing required treatment group(s): ", paste(missing_treatments, collapse = ", "),
    ". Observed treatment groups are: ", paste(observed_treatments, collapse = ", ")
  )
}

if (any(!is.finite(long$cytotoxicity_effect))) {
  stop("Non-finite cytotoxicity effect values detected. Check the input table for missing or non-numeric values.")
}

long
```

# Summarize single-agent and mixture effects

```{r summarize-effects}
summary_by_dose <- long %>%
  dplyr::group_by(dose, treatment) %>%
  dplyr::summarise(
    mean_effect = mean(cytotoxicity_effect, na.rm = TRUE),
    sd_effect = sd(cytotoxicity_effect, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

summary_wide <- summary_by_dose %>%
  dplyr::select(dose, treatment, mean_effect) %>%
  tidyr::pivot_wider(
    names_from = treatment,
    values_from = mean_effect
  )

missing_effect_columns <- setdiff(required_treatments, names(summary_wide))
if (length(missing_effect_columns) > 0) {
  stop(
    "Could not create all required effect columns after pivoting: ",
    paste(missing_effect_columns, collapse = ", ")
  )
}

bliss_summary <- summary_wide %>%
  dplyr::mutate(
    bliss_expected_effect = cd + cr - (cd * cr),
    mix_observed_effect = mix,
    mean_bliss_excess = mix_observed_effect - bliss_expected_effect
  ) %>%
  dplyr::arrange(dose)

bliss_summary
```

# Bootstrap Bliss excess confidence intervals

```{r bootstrap-function}
bootstrap_bliss_excess <- function(data, n_boot = 5000, seed = 831) {
  set.seed(seed)

  doses <- sort(unique(data$dose))

  purrr::map_dfr(doses, function(current_dose) {
    dose_data <- data %>%
      dplyr::filter(dose == current_dose)

    cd_values <- dose_data %>%
      dplyr::filter(treatment == "cd") %>%
      dplyr::pull(cytotoxicity_effect)

    cr_values <- dose_data %>%
      dplyr::filter(treatment == "cr") %>%
      dplyr::pull(cytotoxicity_effect)

    mix_values <- dose_data %>%
      dplyr::filter(treatment == "mix") %>%
      dplyr::pull(cytotoxicity_effect)

    if (length(cd_values) == 0 || length(cr_values) == 0 || length(mix_values) == 0) {
      stop("Dose ", current_dose, " is missing cd, cr, or mix replicate values.")
    }

    boot_excess <- replicate(n_boot, {
      cd_mean <- mean(sample(cd_values, size = length(cd_values), replace = TRUE), na.rm = TRUE)
      cr_mean <- mean(sample(cr_values, size = length(cr_values), replace = TRUE), na.rm = TRUE)
      mix_mean <- mean(sample(mix_values, size = length(mix_values), replace = TRUE), na.rm = TRUE)

      bliss_expected <- cd_mean + cr_mean - (cd_mean * cr_mean)
      mix_mean - bliss_expected
    })

    tibble::tibble(
      dose = current_dose,
      bootstrap_mean_bliss_excess = mean(boot_excess, na.rm = TRUE),
      ci_lower = as.numeric(stats::quantile(boot_excess, probs = 0.025, na.rm = TRUE)),
      ci_upper = as.numeric(stats::quantile(boot_excess, probs = 0.975, na.rm = TRUE)),
      n_boot = n_boot
    )
  })
}
```

```{r bootstrap-analysis}
bliss_ci <- bootstrap_bliss_excess(
  data = long,
  n_boot = n_boot,
  seed = seed
)

bliss_ci
```

# Final Bliss independence results table

```{r final-table}
raw_data <- long %>%
  dplyr::transmute(
    Concentration = dose,
    Condition = treatment,
    Replicate = replicate,
    Normalized_LD_Input = normalized_ld_input,
    Normalized_LD_Ratio = normalized_ld_ratio,
    Cytotoxicity_Effect = cytotoxicity_effect
  ) %>%
  dplyr::arrange(Concentration, Condition, Replicate)

bliss_independence_results <- bliss_summary %>%
  dplyr::select(
    dose,
    cd,
    cr,
    bliss_expected_effect,
    mix_observed_effect,
    mean_bliss_excess
  ) %>%
  dplyr::left_join(bliss_ci, by = "dose") %>%
  dplyr::mutate(
    interaction_type = dplyr::case_when(
      ci_upper < 0 ~ "less than additive (antagonistic)",
      ci_lower > 0 ~ "greater than additive (synergistic)",
      ci_lower <= 0 & ci_upper >= 0 ~ "additive",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::rename(
    Concentration = dose,
    Cd_Effect = cd,
    Cr_Effect = cr,
    Bliss_Expected_Effect = bliss_expected_effect,
    Mix_Observed_Effect = mix_observed_effect,
    Mean_Bliss_Excess = mean_bliss_excess,
    Bootstrap_Mean_Bliss_Excess = bootstrap_mean_bliss_excess,
    CI_Lower = ci_lower,
    CI_Upper = ci_upper,
    N_Bootstrap = n_boot,
    Interaction_Type = interaction_type
  ) %>%
  dplyr::arrange(Concentration)

bliss_independence_results
```

# Visualization

```{r plot-bliss-excess}
bliss_plot <- ggplot2::ggplot(
  bliss_independence_results,
  ggplot2::aes(x = Concentration, y = Mean_Bliss_Excess)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = CI_Lower, ymax = CI_Upper),
    width = 0.08
  ) +
  ggplot2::labs(
    title = "Bliss Independence Analysis of Cd + Cr(VI) Mixture Effects",
    x = "Concentration",
    y = "Bliss excess\nObserved mixture effect - Bliss expected effect"
  ) +
  ggplot2::theme_classic()

bliss_plot
```

# Export results

```{r export-results}
readr::write_csv(
  raw_data,
  file.path(output_dir, "bliss_replicate_level_data.csv")
)

readr::write_csv(
  bliss_independence_results,
  file.path(output_dir, "bliss_independence_results.csv")
)

ggplot2::ggsave(
  filename = file.path(results_dir, "bliss_excess_bootstrap_ci.pdf"),
  plot = bliss_plot,
  width = 7,
  height = 4.5
)

writexl::write_xlsx(
  x = list(
    Raw_Data = raw_data,
    Bliss_Independence_Results = bliss_independence_results
  ),
  path = output_excel
)

cat("Exported files:\n")
cat("- ", file.path(output_dir, "bliss_replicate_level_data.csv"), "\n", sep = "")
cat("- ", file.path(output_dir, "bliss_independence_results.csv"), "\n", sep = "")
cat("- ", file.path(results_dir, "bliss_excess_bootstrap_ci.pdf"), "\n", sep = "")
cat("- ", output_excel, "\n", sep = "")
```

# Session information

```{r session-info}
sessionInfo()
```
